---
title: "Phylogenetic_strain_selection"
author: "KiseokUchicago"
date: "2021-06-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=11, fig.height=9,
                      error=TRUE, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE)
```

## Phylogenetic tree construction
**Kiseok Lee** 


```{r}
# libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(tidyverse)
library(magrittr)
library(readxl)
library(reshape2)
library(gtools)
library(devtools)
library(openxlsx)
library(ape)

## theme for ggplot
mytheme <- theme_bw() + 
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=10, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),panel.border=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))
  
mytheme_2d <- theme_bw() + 
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=13, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))


# color collection
my_color_collection <- c(
  "#CBD588", "#5F7FC7", "orange", "#AD6F3B", "#673770", 
  "#D14285", "#652926", "#C84248", "#8569D5", "#5E738F",
  "#D1A33D", "#8A7C64", "#599861","#616163", "#FFCDB2",
  "#6D9F71", "#242F40",
   "#CCA43B", "#F92A82", "#ED7B84", "#7EB77F", 
  "#DEC4A1", "#E5D1D0", '#0E8482', '#C9DAEA', '#337357', 
  '#95C623', '#E55812', '#04471C', '#F2D7EE', '#D3BCC0', 
  '#A5668B', '#69306D', '#0E103D', '#1A535C', '#4ECDC4', 
  '#F7FFF7', '#FF6B6B', '#FFE66D', '#6699CC', '#FFF275', 
  '#FF8C42', '#FF3C38', '#A23E48', '#000000', '#CF5C36', 
  '#EEE5E9', '#7C7C7C', '#EFC88B', '#2E5266', '#6E8898', 
  '#9FB1BC', '#D3D0CB', '#E2C044', '#5BC0EB', '#FDE74C', 
  '#9BC53D', '#E55934', '#FA7921', "#CD9BCD", "#508578", "#DA5724")


```

## 1. Data input and pre-processing for phylogenetic tree construction
```{r}
df_Biobank <- read.csv('data/biobank.umap.16sSeqs.csv')

dim(df_Biobank)
colnames(df_Biobank)
head(df_Biobank)


# how many species
df_species <- df_Biobank %>% group_by(species) %>% summarise(number=n_distinct(species))
dim(df_species)
head(df_species)

# get id and sequence
# id is seq_id + species
phylo_Biobank <- df_Biobank %>% unite(phy_id,c("seq_id",'species'),sep="_", remove=F)
phylo_Biobank$species <- gsub("\\[","",phylo_Biobank$species)
phylo_Biobank$species <- gsub("]","",phylo_Biobank$species)
colnames(phylo_Biobank)
# remove space in the id for downstream
phylo_Biobank$phy_id <- gsub(" ","_",phylo_Biobank$phy_id) 
# remove "[" and "]"
phylo_Biobank$phy_id <- gsub("\\[","",phylo_Biobank$phy_id)
phylo_Biobank$phy_id <- gsub("]","",phylo_Biobank$phy_id)
col2_Biobank <- phylo_Biobank %>% dplyr::select(phy_id,nuc_sequence)

dim(col2_Biobank)
head(col2_Biobank)

write.table(col2_Biobank, file='input_for_fasta_conversion.txt', quote=FALSE, sep='\t', row.names = F, col.names=F)

```

## 2. UMAP plotting

```{r}
# UMAP plotting
ggplot(df_Biobank, aes(x=umapA1, y=umapA2)) +
  xlab('\n umap A1')+
  ylab("umap A2 \n") +
  geom_point(aes(fill = genus), shape = 21 ,size=4, alpha=0.9) +
  scale_fill_manual(values = my_color_collection)+
  ggtitle(paste0("UMAP coordination"," \n")) +
  # scale_x_continuous(limits = c(0, max(df_cen_wb$Degree))) +
  # scale_y_continuous(limits = c(0, max(df_cen_wb$Betweenness))) +
  theme(legend.text=element_text(size=13)) + 
  # theme(legend.position="top") +
  # theme(legend.title=element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=8), reverse = TRUE))+
  guides(size=FALSE) +
  mytheme_2d

plot_UMAP <- function(df_Biobank){
  ggplot(df_Biobank, aes(x=umapA1, y=umapA2)) +
    xlab('\n umap A1')+
    ylab("umap A2 \n") +
    geom_point(aes(fill = genus), shape = 21 ,size=4, alpha=0.9) +
    scale_fill_manual(values = my_color_collection)+
    ggtitle(paste0("UMAP coordination"," \n")) +
    # scale_x_continuous(limits = c(0, max(df_cen_wb$Degree))) +
    # scale_y_continuous(limits = c(0, max(df_cen_wb$Betweenness))) +
    theme(legend.text=element_text(size=13)) + 
    # theme(legend.position="top") +
    # theme(legend.title=element_blank()) +
    guides(colour = guide_legend(override.aes = list(size=8), reverse = TRUE))+
    guides(size=FALSE) +
    mytheme_2d
}


```

## 2.2. K-means clustering for UMAP distribution

(1) First, using the elbow method to find the optimal number of clusters \
Within-Cluster-Sum-of-Squares (WCSS) is the sum of squares of the distances of each data point in all clusters to their respective centroids. \

(2) 
```{r}
# First, using the elbow method to find the optimal number of clusters
set.seed(6)
wcss = vector()

colnames(df_Biobank)
k_dataset <- phylo_Biobank %>% dplyr::select(phy_id, umapA1, umapA2)
k_dataset <- tibble::column_to_rownames(k_dataset, var="phy_id")

# 100 clusters
n_clu = 100
for (i in 1:n_clu) wcss[i] = sum(kmeans(k_dataset, i)$withinss)
plot(1:n_clu,
     wcss,
     type = 'b',
     main = paste('The Elbow Method'),
     xlab = 'Number of clusters',
     ylab = 'WCSS')

# Fitting K-Means to the dataset
set.seed(29)
kmeans = kmeans(x = k_dataset, centers = n_clu)
y_kmeans = kmeans$cluster

# Visualising the clusters
library(cluster)
# All observation are represented by points in the plot, using principal components or multidimensional scaling.
clusplot(k_dataset,
         y_kmeans,
         lines = 0,
         shade = TRUE,
         color = TRUE,
         labels = 0,
         plotchar = FALSE,
         span = TRUE,
         # main = paste('Clusters of strains with UMAP'),
         main = NA,
         xlab = 'PC1',
         ylab = 'PC2') 


# get the centroid and the closest id's for 100 clusters
kmeans$centers

center_strains <- vector()
min_ss <- rep(Inf, n_clu)  # minimum sum of square of distance from centroid

for (i in 1:n_clu){
  print(i)
  centroid_xy <- (kmeans$centers)[i,]
  for (j in 1:dim(k_dataset)[1]){
    point_xy <- k_dataset[j,]
    ss <- sum((centroid_xy - point_xy)^2)
    if (ss < min_ss[i]){
      min_ss[i] <- ss
      center_strains[i] <- rownames(k_dataset[j,])
      # print(center_strains[i])
    }
  }
}

kmeans$cluster[center_strains] # center strains belong to each cluster
center_strains # confirmed
length(center_strains)

# Plot UMAP with cluster center strains
plot_UMAP(phylo_Biobank %>% filter(phy_id %in% center_strains))

# number of species
df_center_umap <- phylo_Biobank %>% filter(phy_id %in% center_strains)
df_center_umap %>% dplyr::select(species) %>% unique() %>% dim() # 51 species
df_umap_distinct <- df_center_umap %>% count(species)
hist(df_umap_distinct$n, main="Histogram of number of same species", xlab="Number of same species")

```

Try it again with 20 clusters
```{r}
# 20 clusters
set.seed(6)
wcss = vector()
n_clu = 20
for (i in 1:n_clu) wcss[i] = sum(kmeans(k_dataset, i)$withinss)
plot(1:n_clu,
     wcss,
     type = 'b',
     main = paste('The Elbow Method'),
     xlab = 'Number of clusters',
     ylab = 'WCSS')

# Fitting K-Means to the dataset
set.seed(29)
kmeans = kmeans(x = k_dataset, centers = n_clu)
y_kmeans = kmeans$cluster

# Visualising the clusters
library(cluster)
# All observation are represented by points in the plot, using principal components or multidimensional scaling.
clusplot(k_dataset,
         y_kmeans,
         lines = 0,
         shade = TRUE,
         color = TRUE,
         labels = 0,
         plotchar = FALSE,
         span = TRUE,
         # main = paste('Clusters of strains with UMAP'),
         main = NA,
         xlab = 'PC1',
         ylab = 'PC2') 


# get the centroid and the closest id's for 100 clusters
kmeans$centers

center_strains <- vector()
min_ss <- rep(Inf, n_clu)  # minimum sum of square of distance from centroid

for (i in 1:n_clu){
  print(i)
  centroid_xy <- (kmeans$centers)[i,]
  for (j in 1:dim(k_dataset)[1]){
    point_xy <- k_dataset[j,]
    ss <- sum((centroid_xy - point_xy)^2)
    if (ss < min_ss[i]){
      min_ss[i] <- ss
      center_strains[i] <- rownames(k_dataset[j,])
      # print(center_strains[i])
    }
  }
}

kmeans$cluster[center_strains] # center strains belong to each cluster
center_strains # confirmed
length(center_strains)

# Plot UMAP with cluster center strains
plot_UMAP(phylo_Biobank %>% filter(phy_id %in% center_strains))

```


## 3. Phylogenetic tree construction and cut the tree / clustering

```{r}
phy_tree <- ape::read.tree("data/Biobank_align_mafft.fasta.treefile_rooted.nwk")
phy_tree

# will be using cutree function in hclust package. In order to do so, we need to make it ultrametric
# not ultrametric
is.ultrametric(phy_tree)

# convert to ultrametric tree 
# (http://blog.phytools.org/2017/03/forceultrametric-method-for-ultrametric.html)
force.ultrametric<-function(tree,method=c("nnls","extend")){
    method<-method[1]
    if(method=="nnls") tree<-nnls.tree(cophenetic(tree),tree,
        rooted=TRUE,trace=0)
    else if(method=="extend"){
        h<-diag(vcv(tree))
        d<-max(h)-h
        ii<-sapply(1:Ntip(tree),function(x,y) which(y==x),
            y=tree$edge[,2])
        tree$edge.length[ii]<-tree$edge.length[ii]+d
    } else 
        cat("method not recognized: returning input tree\n\n")
    tree
}

library(phangorn)
ult_tree <- force.ultrametric(phy_tree)
is.ultrametric(ult_tree) # true

# convert to hclust object
hcl <- as.hclust(ult_tree)
plot(hcl,labels = FALSE, hang= -1)

```

The cutree() function provides the functionality to output either desired number of clusters or clusters obtained from cutting the dendrogram at a certain height.

```{r}
# cut tree
cut_tree <- cutree(hcl, k=100)
table(cut_tree) 
hist(as.vector(table(cut_tree)), breaks=100)

# table
table(cut_tree)
head(cut_tree)

# this is not a good method to cluster
```

## 3.2. Use distance matrix to do k-means clustering

```{r}
# input tab delimited table
ml_dist <- read.table("data/Biobank_align_mafft.fasta.mldist.tsv", sep="\t", header = F)
# head(ml_dist)

# set row and column name
ml_dist %<>% dplyr::select(-V531)
dim(ml_dist)

colnames(ml_dist)
tibble::column_to_rownames(ml_dist, var="V1") %>% dim()

mat_dist <- tibble::column_to_rownames(ml_dist, var="V1")
colnames(mat_dist) <- rownames(mat_dist)
dim(mat_dist)

mat_dist <- as.matrix(mat_dist)

```

K-means clustering

```{r}
# distance matrix has 529 x 529 dimension
library(FCPS)

kmeans_100 <- kmeansClustering(mat_dist, ClusterNo=100, RandomNo=1, PlotIt= T, Verbose = T)

kmeans_cls <- kmeans_100$Cls
table(kmeans_cls)

# Histogram of # of strains belonging to each cluster
hist(as.vector(table(kmeans_cls)), breaks=100)

# Let's get the list of center ids
# kmeans_100

kmeans_100$centerids
phy_id_list <- names(kmeans_cls[kmeans_100$centerids])

# function to get the centroid ids
get_centroids <- function(mat_dist, num_cluster=100){
  kmeans_100 <- kmeansClustering(mat_dist, ClusterNo=num_cluster, RandomNo=1, PlotIt= T, Verbose = T)
  kmeans_cls <- kmeans_100$Cls
  table(kmeans_cls)
  
  # Let's get the list of center ids
  phy_id_list <- names(kmeans_cls[kmeans_100$centerids])
  return(phy_id_list)
}

# get 100 centroids and plot UMAP
phy_id_list <- get_centroids(mat_dist, num_cluster=100)

df_centroid_100 <- phylo_Biobank %>% filter(phy_id %in% phy_id_list)

plot_UMAP(phylo_Biobank %>% filter(phy_id %in% phy_id_list))

# number of species
df_center_phy <- phylo_Biobank %>% filter(phy_id %in% phy_id_list)
df_center_phy %>% dplyr::select(species) %>% unique() %>% dim() # 47 species
df_phy_distinct <- df_center_phy %>% count(species)
hist(df_phy_distinct$n, main="Histogram of number of same species", xlab="Number of same species")

# See what species the centerids are
phylo_Biobank %>% filter(phy_id %in% phy_id_list) %>% dplyr::select(phy_id, species) %>% arrange(species)

# looking into the clusters
kmeans_cls[kmeans_cls==90]

# searching
df_cls <- data.frame(kmeans_cls)
df_cls <- tibble::rownames_to_column(df_cls, var="id")
colnames(df_cls)


df_cls %>% filter(str_detect(id, "Bacteroides_fragilis"))


```

## 4. Comparison of 100 strain sets
Now we will calculate NRI (Net Relatedness Index) and NTI (Nearest Taxon Index) for our different communities. Rembember, negative NRI and NTI values indicate a high level of phylogenetic overdispersion, and positive NRI and NTI values indicate phylogenetic clustering. \
- Source: http://ib.berkeley.edu/courses/ib200/2016/labs/14/lab14.R \
- Source2: http://ib.berkeley.edu/courses/ib200b/labs/lab16/ib200b_lab16_phylocomm.pdf \
- Paper: Horner-Devine, M. C., & Bohannan, B. J. (2006). Phylogenetic clustering and overdispersion in bacterial communities. Ecology, 87(sp7), S100-S108. https://esajournals.onlinelibrary.wiley.com/doi/10.1890/0012-9658%282006%2987%5B100%3APCAOIB%5D2.0.CO%3B2 \

```{r}
# PICANTE R package
# install.packages("picante")
library(picante)

# strain list for 2 different methods
# (Method 1) UMAP method
strains_umap <- center_strains
# (Method 2) Phylogenetic tree method
strains_phy <- phy_id_list

# Let's visualize our data. Now let's see how taxa from the 2 communities from different methods arranged on the tree. 
phy_tree
phy_tree$edge

n_tips <- length(phy_tree$tip.label)

# let's color the branches
df_tree_color <- tibble(phy_id = phy_tree$tip.label, color=rep(1,n_tips),
       umap = ifelse(phy_tree$tip.label %in% strains_umap,T,F),
       phy = ifelse(phy_tree$tip.label %in% strains_phy,T,F))


sum(df_tree_color$umap) # confirmed 100
sum(df_tree_color$phy) # confirmed 100

# color
df_tree_color$color <- ifelse(df_tree_color$umap & df_tree_color$phy,4,# if in both umap and phy (blue)
                              ifelse(df_tree_color$phy, 3, # if in only phy (green)
                                     ifelse(df_tree_color$umap, 2, 1))) # if in only umap (red)
table(df_tree_color$color) # 1: black, 2: umap, 3: phy, 4: both

# (1) Plot tree with umap-method strains indicated
plot.phylo(phy_tree, show.tip.label = F, cex=0.1,
           # align.tip.label = 2,
           # tip.color=df_tree_color$umap, 
           edge.width=1,
           # adj = c(1, 0.5),
           no.margin = T)
tiplabels(tip = which(df_tree_color$umap > 0), pch = 19, cex = 2, col='maroon')

# (2) Plot tree with umap-method strains indicated
plot.phylo(phy_tree, show.tip.label = F, cex=0.1,
           # align.tip.label = 2,
           # tip.color=df_tree_color$umap, 
           edge.width=1,
           # adj = c(1, 0.5),
           no.margin = T)
tiplabels(tip = which(df_tree_color$phy > 0), pch = 19, cex = 2, col='skyblue')

# plot together with colors
plot.phylo(phy_tree, show.tip.label = F, cex=0.1,
           # align.tip.label = 2,
           # tip.color=df_tree_color$umap, 
           edge.width=1,
           # adj = c(1, 0.5),
           no.margin = T)
tiplabels(tip = which(df_tree_color$color == 2), pch = 19, cex = 2, col='maroon')
tiplabels(tip = which(df_tree_color$color == 3), pch = 19, cex = 2, col='skyblue')
tiplabels(tip = which(df_tree_color$color == 4), pch = 19, cex = 2, col='navy')

```

Calculate NRI and NTI

```{r}
# making binary community matrix
umap = phy_tree$tip.label %in% strains_umap
phylogenetic = phy_tree$tip.label %in% strains_phy

# get community matrix with different species
select_different_species <- function(n=100){
  phy_species <- phylo_Biobank %>% select(phy_id, species)
  unique_species_vec <- unique(phylo_Biobank$species)  # unique species vector
  
  # randomly select n=100 different species
  diff_species_rand1 <- sample(unique_species_vec, n)
  phy_species %<>% filter(species %in% diff_species_rand1) # get strains with those sampled species
  dim(phy_species)
  
  # randomly select 1 strain of each 100 species
  diff_species <- phy_species[sample(1:dim(phy_species)[1]),] %>% group_by(species) %>% filter(row_number()==1)
  length(unique(diff_species$species)) # 100

  diff_species1 = phy_tree$tip.label %in% diff_species$phy_id
  return(diff_species1)
}

select_different_species(n=100)

set.seed(1)
dif_species1 = select_different_species(n=100)
dif_species2 = select_different_species(n=100)
dif_species3 = select_different_species(n=100)

# random 
set.seed(61)
random1 = phy_tree$tip.label %in% sample(phy_tree$tip.label, 100)
random2 = phy_tree$tip.label %in% sample(phy_tree$tip.label, 100)
random3 = phy_tree$tip.label %in% sample(phy_tree$tip.label, 100)

# bind it into community matrix
mat_com <- rbind(umap, phylogenetic, dif_species1, dif_species2, dif_species3, random1, random2, random3)
colnames(mat_com) <- phy_tree$tip.label

dim(mat_com)
phydist <- cophenetic(phy_tree)
dim(phydist)

ses.mpd(mat_com, dis = phydist, null.model="taxa.labels")

NRI_table <- ses.mpd(mat_com, dis = phydist, null.model="taxa.labels")
NRI_table <- tibble::rownames_to_column(NRI_table, var="method")
write.xlsx(NRI_table, "NRI_table.xlsx")

# The rows are the communities.  The first four columns should be pretty straight forward given the definition of the NRI from lecture.  To review:
# ntaxa Number of taxa in community
# mpd.obs Observed mean pairwise distance (MPD) in community
# mpd.rand.mean Mean MPD in null communities
# mpd.rand.sd Standard deviation of MPD in null communities
# mpd.obs.rank Rank of observed MPD vs. null communities
# mpd.obs.z Standardized effect size of MPD vs. null communities (equivalent to -NRI)

NTI_table <- ses.mntd(mat_com, dis = phydist, null.model="taxa.labels")
NTI_table <- tibble::rownames_to_column(NTI_table, var="method")
write.xlsx(NTI_table, "NTI_table.xlsx")
```

Histogram of species
```{r}

phy_species <- phylo_Biobank %>% select(phy_id, species)
hist_spe <- phy_species %>% group_by(species) %>% summarize(frequency=n_distinct(phy_id))
histogram(hist_spe$frequency, xlab="Number of Strains in the same species", breaks=30)

```

## 5. Selection of strains
- Selecting strains belonging to different species is the optimal strategy.
- Randomly pick and choose the community with lowest NTI.

```{r}
# matrix
n <- 1000
n_tips <- length(phy_tree$tip.label)

mat_1000 <- matrix(rep(-1,n*n_tips), nrow=n,ncol=n_tips)
dim(mat_1000)
for (i in 1:n){
  mat_1000[i,] = select_different_species(n=95)
}

colnames(mat_1000) <- phy_tree$tip.label
rownames(mat_1000) <- paste0("set",1:n)
# head(mat_1000)

NTI_species_table <- ses.mntd(mat_1000, dis = phydist, null.model = "taxa.labels")
NTI_species_table <- tibble::rownames_to_column(NTI_species_table, var="method")
NTI_species_table %>% filter(mntd.obs.z == max(NTI_species_table$mntd.obs.z))

set_number <- NTI_species_table %>% filter(mntd.obs.z == max(NTI_species_table$mntd.obs.z))
set_number <- set_number$method
set_number

selected_strain_list <- mat_1000[set_number,]
selected_strain_vec <- names(selected_strain_list[selected_strain_list > 0])

# get excel
colnames(phylo_Biobank)
df_95strains <- phylo_Biobank %>% filter(phy_id %in% selected_strain_vec)
write.xlsx(df_95strains,"Strain_selected_95_max_NTI.xlsx")

# plot UMAP
plot_UMAP(phylo_Biobank %>% filter(phy_id %in% selected_strain_vec))

# plot tree
df_tree_color <- tibble(phy_id = phy_tree$tip.label, species = ifelse(phy_tree$tip.label %in% selected_strain_vec,T,F))

plot.phylo(phy_tree, show.tip.label = F, cex=0.1,
           # align.tip.label = 2,
           # tip.color=df_tree_color$umap, 
           edge.width=1,
           # adj = c(1, 0.5),
           no.margin = T)
tiplabels(tip = which(df_tree_color$species > 0), pch = 19, cex = 2, col='green')


```





