---
title: "select20_from_66strains"
author: "KiseokUchicago"
date: "2021-07-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=11, fig.height=9,
                      error=TRUE, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE)
```

## Pilot experiment: 20 strain selection from 66 strain constraint
**Kiseok Lee** 
PI: **Seppe Kuehn**, **Arjun Raman**


```{r}
# libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(tidyverse)
library(magrittr)
library(readxl)
library(reshape2)
library(gtools)
library(devtools)
library(openxlsx)
library(ape)
library(picante)
library(fields)

## theme for ggplot
mytheme <- theme_bw() + 
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=10, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),panel.border=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))
  
mytheme_2d <- theme_bw() + 
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=13, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))


# color collection
my_color_collection <- c(
  "#CBD588", "#5F7FC7", "orange", "#AD6F3B", "#673770", 
  "#D14285", "#652926", "#C84248", "#8569D5", "#5E738F",
  "#D1A33D", "#8A7C64", "#599861","#616163", "#FFCDB2",
  "#6D9F71", "#242F40",
   "#CCA43B", "#F92A82", "#ED7B84", "#7EB77F", 
  "#DEC4A1", "#E5D1D0", '#0E8482', '#C9DAEA', '#337357', 
  '#95C623', '#E55812', '#04471C', '#F2D7EE', '#D3BCC0', 
  '#A5668B', '#69306D', '#0E103D', '#1A535C', '#4ECDC4', 
  '#F7FFF7', '#FF6B6B', '#FFE66D', '#6699CC', '#FFF275', 
  '#FF8C42', '#FF3C38', '#A23E48', '#000000', '#CF5C36', 
  '#EEE5E9', '#7C7C7C', '#EFC88B', '#2E5266', '#6E8898', 
  '#9FB1BC', '#D3D0CB', '#E2C044', '#5BC0EB', '#FDE74C', 
  '#9BC53D', '#E55934', '#FA7921', "#CD9BCD", "#508578", "#DA5724")

# functions
plot_UMAP <- function(df_Biobank){
  ggplot(df_Biobank, aes(x=umapA1, y=umapA2)) +
    xlab('\n umap A1')+
    ylab("umap A2 \n") +
    geom_point(aes(fill = genus), shape = 21 ,size=4, alpha=0.9) +
    scale_fill_manual(values = my_color_collection)+
    ggtitle(paste0("UMAP coordination"," \n")) +
    # scale_x_continuous(limits = c(0, max(df_cen_wb$Degree))) +
    # scale_y_continuous(limits = c(0, max(df_cen_wb$Betweenness))) +
    theme(legend.text=element_text(size=13)) + 
    # theme(legend.position="top") +
    # theme(legend.title=element_blank()) +
    guides(colour = guide_legend(override.aes = list(size=8), reverse = TRUE))+
    guides(size=FALSE) +
    mytheme_2d
}

# import tree
phy_tree <- ape::read.tree("data/Biobank_align_mafft.fasta.treefile_rooted.nwk")
phy_tree

# import database
df_Biobank <- read.csv('data/biobank.umap.16sSeqs.csv')
# how many species
df_species <- df_Biobank %>% group_by(species) %>% summarise(number=n_distinct(species))
# get id and sequence
# id is seq_id + species
phylo_Biobank <- df_Biobank %>% unite(phy_id,c("seq_id",'species'),sep="_", remove=F)
phylo_Biobank$species <- gsub("\\[","",phylo_Biobank$species)
phylo_Biobank$species <- gsub("]","",phylo_Biobank$species)
# remove space in the id for downstream
phylo_Biobank$phy_id <- gsub(" ","_",phylo_Biobank$phy_id) 
# remove "[" and "]"
phylo_Biobank$phy_id <- gsub("\\[","",phylo_Biobank$phy_id)
phylo_Biobank$phy_id <- gsub("]","",phylo_Biobank$phy_id)
col2_Biobank <- phylo_Biobank %>% dplyr::select(phy_id,nuc_sequence)
```

## 1. Import 66 strain list 
This list is achieved from filtering out strains from the 95 strain list that may not grow in BHIS media and are not strictly anaerobic. 

```{r}
strain66 <- read.xlsx('data/Strain_selected_95_max_NTI_CEKannotated_updated.xlsx')
dim(strain66)
head(strain66)
colnames(strain66)
strain66 <- strain66 %>% filter(!is.na(phy_id))

# Plot UMAP
plot_UMAP(strain66)

# Plot tree
strain66_vec = strain66$phy_id
df_tree_color <- tibble(phy_id = phy_tree$tip.label, species = ifelse(phy_tree$tip.label %in% strain66_vec,T,F))

plot.phylo(phy_tree, show.tip.label = F, cex=0.1,
           # align.tip.label = 2,
           # tip.color=df_tree_color$umap, 
           edge.width=1,
           # adj = c(1, 0.5),
           no.margin = T)
tiplabels(tip = which(df_tree_color$species > 0), pch = 19, cex = 2, col='darkgreen')


```


## 2. Strategy
1. Maximizing mean nearest UMAP euclidean distance (2D)
- nearest UMAP distance for 1 strain = UMAP distance to nearest strain 
- mean nearest UMAP distance for 1 strain = average of UMAP distance to nearest strain for all strain

```{r}
# Get 20 strain set with highest "mean nearest UMAP distance"

# randomly select n=20 different species
rand_num = 10000 # how many sampling 
n = 20 # community size
strain20_rand <- sample(strain66_vec, n)

library(fields)
strain66_coor <- strain66 %>% select(phy_id, umapA1, umapA2)
strain66_coor <- tibble::column_to_rownames(strain66_coor, var="phy_id") 
strain66_dist <- rdist(strain66_coor)
dim(strain66_dist)
strain66_dist <- as.data.frame(strain66_dist)
rownames(strain66_dist) <- rownames(strain66_coor)
colnames(strain66_dist) <- rownames(strain66_coor)
head(strain66_dist)

# get minimum distance (minimum positive)
apply(strain66_dist, 1, FUN = function(x) {min(x[x > 0])})
# get mean of nearest UMAP distance
mean(apply(strain66_dist, 1, FUN = function(x) {min(x[x > 0])}))

# execute
sampling20_list <- list() 

for (rand in 1:rand_num){
  print(rand)
  # random sampling
  strain20_rand <- sample(strain66_vec, n)
  strain20_coor <- strain66 %>% filter(phy_id %in% strain20_rand) %>% select(phy_id, umapA1, umapA2)
  strain20_coor <- tibble::column_to_rownames(strain20_coor, var="phy_id") 
  strain20_dist <- rdist(strain20_coor)
  dim(strain20_dist)
  strain20_dist <- as.data.frame(strain20_dist)
  rownames(strain20_dist) <- rownames(strain20_coor)
  colnames(strain20_dist) <- rownames(strain20_coor)
  head(strain20_dist)
  
  # get minimum distance (minimum positive)
  # apply(strain66_dist, 1, FUN = function(x) {min(x[x > 0])})
  # get mean of nearest UMAP distance
  mean1 = mean(apply(strain20_dist, 1, FUN = function(x) {min(x[x > 0])}))
  
  # save in list
  sampling20_list[[rand]] <- list(
    mean_nearest_distance = mean1,
    distance_mat = strain20_dist
  )
}

mean_nearest_dist_vec <- c()
for (i in 1:rand_num){
  mean_nearest_dist_vec[i] <- sampling20_list[[i]]$mean_nearest_distance
}

hist(mean_nearest_dist_vec)
which.max(mean_nearest_dist_vec)
sampling20_list[[which.max(mean_nearest_dist_vec)]]$distance_mat

max <- sampling20_list[[which.max(mean_nearest_dist_vec)]]$distance_mat
max_strain_list <- rownames(max)

## plot UMAP of selected strains

plot_UMAP(strain66 %>% filter(phy_id %in% max_strain_list))

```

This method does not work well.

## 2.2. Adjust lower percentile to average

```{r}
# randomly select n=20 different species
rand_num = 10000 # how many sampling 
n = 20 # community size

# execute function
sampling_func <- function(rand_num = 10000, n = 20, lower_percentile = 1.0){
  # empty list
  sampling20_list <- list() 
  
  # do sampling for rand_num of times
  for (rand in 1:rand_num){
    # print(rand)
    # random sampling
    strain20_rand <- sample(strain66_vec, n)
    strain20_coor <- strain66 %>% filter(phy_id %in% strain20_rand) %>% select(phy_id, umapA1, umapA2)
    strain20_coor <- tibble::column_to_rownames(strain20_coor, var="phy_id") 
    strain20_dist <- rdist(strain20_coor)
    dim(strain20_dist)
    strain20_dist <- as.data.frame(strain20_dist)
    rownames(strain20_dist) <- rownames(strain20_coor)
    colnames(strain20_dist) <- rownames(strain20_coor)
    head(strain20_dist)
    
    # get minimum distance (minimum positive)
    nearest_dist_vec <- apply(strain20_dist, 1, FUN = function(x) {min(x[x > 0])})
    nearest_dist_vec <- nearest_dist_vec[order(nearest_dist_vec)] # order from lowest to highest
    
    
    # nearest distance for lower percentile
    nearest_dist_vec[1:round(n*lower_percentile)]
    # get mean of nearest UMAP distance
    mean1 = mean(nearest_dist_vec[1:round(n*lower_percentile)])
    
    # save in list
    sampling20_list[[rand]] <- list(
      mean_nearest_distance = mean1,
      distance_mat = strain20_dist
    )
  }
  return(sampling20_list)
}

```

100% percentile

```{r}
strain20_100percentile <- sampling_func(rand_num = 10000, n = 20, lower_percentile = 1.0)

  
mean_nearest_dist_vec <- c()
for (i in 1:rand_num){
  mean_nearest_dist_vec[i] <- strain20_100percentile[[i]]$mean_nearest_distance
}

hist(mean_nearest_dist_vec)
which.max(mean_nearest_dist_vec)
strain20_100percentile[[which.max(mean_nearest_dist_vec)]]$distance_mat

max_100 <- strain20_100percentile[[which.max(mean_nearest_dist_vec)]]$distance_mat
max_100strain_list <- rownames(max_100)

## plot UMAP of selected strains

plot_UMAP(strain66 %>% filter(phy_id %in% max_100strain_list))

strain20_100per <- phylo_Biobank %>% filter(phy_id %in% max_100strain_list)
write.xlsx(strain20_100per,"Strain_selected_20_from_66_ver1.xlsx")

```

average of lower 60% percentile

```{r}
strain20_60percentile <- sampling_func(rand_num = 10000, n = 20, lower_percentile = 0.6)

  
mean_nearest_dist_vec <- c()
for (i in 1:rand_num){
  mean_nearest_dist_vec[i] <- strain20_60percentile[[i]]$mean_nearest_distance
}

hist(mean_nearest_dist_vec)
which.max(mean_nearest_dist_vec)
strain20_60percentile[[which.max(mean_nearest_dist_vec)]]$distance_mat

max_60 <- strain20_60percentile[[which.max(mean_nearest_dist_vec)]]$distance_mat
max_60strain_list <- rownames(max_60)

## plot UMAP of selected strains

plot_UMAP(strain66 %>% filter(phy_id %in% max_60strain_list))

strain20_60per <- phylo_Biobank %>% filter(phy_id %in% max_60strain_list)
write.xlsx(strain20_60per,"Strain_selected_20_from_66_ver2.xlsx")

```

average of lower 30% percentile

```{r}
strain20_30percentile <- sampling_func(rand_num = 10000, n = 20, lower_percentile = 0.3)

  
mean_nearest_dist_vec <- c()
for (i in 1:rand_num){
  mean_nearest_dist_vec[i] <- strain20_30percentile[[i]]$mean_nearest_distance
}

hist(mean_nearest_dist_vec)
which.max(mean_nearest_dist_vec)
strain20_30percentile[[which.max(mean_nearest_dist_vec)]]$distance_mat

max_30 <- strain20_30percentile[[which.max(mean_nearest_dist_vec)]]$distance_mat
max_30strain_list <- rownames(max_30)

## plot UMAP of selected strains

plot_UMAP(strain66 %>% filter(phy_id %in% max_30strain_list))

strain20_30per <- phylo_Biobank %>% filter(phy_id %in% max_30strain_list)
write.xlsx(strain20_30per,"Strain_selected_20_from_66_ver3.xlsx")

```













